Optimal Pricing and Informal Sharing: \\ Evidence from Piped Water in Manila

### 2019-02-01 phil_paper/paper_wjv.tex
restud

%\usepackage[notablist,nofiglist]{endfloat}

@@

%\usepackage[dvipsnames]{xcolor}
%\usepackage{cite}
\onehalfspacing
\usepackage{natbib}  
%\title{ Optimal Pricing and Informal Sharing: \\ Evidence from Piped Water in Manila }
\title{OPTIMAL PRICING AND INFORMAL SHARING: \\ EVIDENCE FROM PIPED WATER IN MANILA}

@@

Faced with high fixed costs of purchasing and maintaining durable goods, consumers often share these goods with each other.  While online platforms have accelerated peer-to-peer rentals of cars, apartments, and storage space (\cite{horton2016owning}), households in developing countries often share connections to public utilities like water and electricity with their neighbors.\footnote{70\% of African water utilities report frequent sharing of water taps between households.  Over 14\% of urban households in Arba-Minch, Ethiopia report sourcing electricity from their neighbors instead of purchasing directly from the utility (\cite{mustefa2016energy}).  In South Africa as of 2012, 11.9\% of the population report sharing their mobile phone with others.  This percentage jumps to 19.9\% among those living in poverty (\cite{calandro2012mobile}).}  Sharing can have large consequences for optimal pricing, particularly for public utilities.  This paper focuses empirically on pricing for piped water in Manila, Philippines in order to increase access and welfare when households share connections.

Improving access to piped water can generate substantial health benefits\footnote{\cite{zhang2016long,kosec2014child,gamper2010impact}, and \cite{galiani2005water} focus on health improvements in developing countries while \cite{cutler2005role} look historically at the US.} as well as improvements in household well-being through time savings and reduced intra-household conflict.\footnote{\cite{devoto2012happiness} provide water through a randomized-controlled trial in Morocco finding few health effects but large benefits to household welfare.}  Yet, the share of urban households without piped water has remained stable at 5\% for middle income countries and at 20\% for low income countries (World Bank [2015]).  While external health and other benefits may merit subsidizing piped water, any subsidies to public utilities are often limited by high costs of public funds and risk of regulatory capture in developing countries (\cite{laffont2005regulation}).  In many cases, prices must be set to fully recover costs.

@@

Insecure land-tenure, credit-constraints, and administrative costs often make it difficult for disadvantaged households to purchase their own water connections (\cite{devoto2012happiness}).  As a result, households frequently share connections including fetching from neighbors with hoses and buckets in Tangiers, Morocco (\cite{devoto2012happiness}) and drawing from single taps that each serve compounds in Kumasi, Ghana (\cite{whittington1992possible}).\footnote{\cite{nauges2006water} finds further evidence of widespread water sharing in Sri Lanka.}  While documenting the prevalence of these  sharing networks, research has not yet examined the importance of water sharing for optimal pricing policy.

>>>>

\usepackage[notablist,nofiglist]{endfloat}

@@

%\usepackage[dvipsnames]{xcolor}
%\usepackage{cite}
\onehalfspacing
\usepackage{natbib}  
\title{ Optimal Pricing and Informal Sharing: \\ Evidence from Piped Water in Manila }
%\title{OPTIMAL PRICING AND INFORMAL SHARING: \\ EVIDENCE FROM PIPED WATER IN MANILA}

@@

%like water and electricity

Faced with high fixed costs of purchasing and maintaining durable goods, consumers often share these goods with each other.  While online platforms have accelerated peer-to-peer rentals of cars, apartments, and storage space (\cite{horton2016owning}), households in developing countries often share connections to public utilities with their neighbors.\footnote{70\% of African water utilities report frequent sharing of water taps between households.  Over 14\% of urban households in Arba-Minch, Ethiopia report sourcing electricity from their neighbors instead of purchasing directly from the utility (\cite{mustefa2016energy}).  In South Africa as of 2012, 11.9\% of the population report sharing their mobile phone with others.  This percentage jumps to 19.9\% among those living in poverty (\cite{calandro2012mobile}).}  Sharing can have large consequences for optimal pricing, particularly for public utilities.  This paper focuses empirically on pricing for piped water when households share connections.  Improving access to piped water can generate substantial health benefits\footnote{\cite{zhang2016long,kosec2014child,gamper2010impact}, and \cite{galiani2005water} focus on health improvements in developing countries while \cite{cutler2005role} look historically at the US.} as well as improvements in household well-being through time savings and reduced intra-household conflict.\footnote{\cite{devoto2012happiness} provide water through a randomized-controlled trial in Morocco finding few health effects but large benefits to household welfare.}  Yet, the share of urban households without piped water has remained stable at 5\% for middle income countries and at 20\% for low income countries (World Bank [2015]).  While external health and other benefits may merit subsidizing piped water, any subsidies to public utilities are often limited by high costs of public funds and risk of regulatory capture in developing countries (\cite{laffont2005regulation}).  In many cases, prices must be set to fully recover costs.

@@

Insecure land-tenure, credit-constraints, and administrative costs often make it difficult for disadvantaged households to purchase their own water connections (\cite{devoto2012happiness}).  As a result, households frequently share connections by fetching from neighbors with hoses and buckets like in Tangiers, Morocco (\cite{devoto2012happiness}) and drawing from single taps that serve many dwellings like in Kumasi, Ghana (\cite{whittington1992possible}).\footnote{\cite{nauges2006water} finds further evidence of widespread water sharing in Sri Lanka.}  While documenting the prevalence of these  sharing networks, research has not yet examined the importance of water sharing for optimal pricing policy.
